---
title: "Understanding and predicting wildfire extremes in the contiguous United States"
subtitle: "https://github.com/mbjoseph/wildfire-extremes"
author: "Max Joseph (@mxwlj), Matt Rossi, Nate Mietkiewicz, Adam Mahood, Megan Cattau, Lise St. Denis, Chelsea Nagy, Virginia Iglesias, John Abatzoglou, Jennifer Balch"
output:
  beamer_presentation:
    fig_caption: false
    latex_engine: xelatex
    fonttheme: "structurebold"
header-includes: 
- \usepackage{graphicx}
- \beamertemplatenavigationsymbolsempty
- \usefonttheme{professionalfonts}
- \renewcommand{\familydefault}{\sfdefault}
- \usepackage{xcolor}
- \definecolor{foreground}{RGB}{0,0,0}
- \definecolor{background}{RGB}{255,255,255}
- \definecolor{title}{RGB}{0,120,255}
- \definecolor{gray}{RGB}{155,155,155}
- \definecolor{subtitle}{RGB}{204,0,204}
- \definecolor{hilight}{RGB}{102,255,204}
- \definecolor{vhilight}{RGB}{255,111,207}
- \setbeamercolor{titlelike}{fg=title}
- \setbeamercolor{subtitle}{fg=subtitle}
- \setbeamercolor{institute}{fg=gray}
- \setbeamercolor{normal text}{fg=foreground,bg=background}
- \setbeamercolor{local structure}{fg=title}
- \setbeamertemplate{frametitle}{\begin{centering} \insertframetitle \par \end{centering}}
---

```{r setup, include=FALSE}
h <- 2.2
knitr::opts_chunk$set(echo = FALSE, 
                      fig.width = 1.618 * h, fig.height = h, 
                      message = FALSE, fig.align = 'center')
library(rgeos)
library(sf)
library(tidyverse)
library(ggthemes)
library(patchwork)
library(RColorBrewer)
library(splines)
theme_set(theme_minimal() + 
            theme(panel.grid.minor = element_blank()))
```

# 

```{r, results='hide'}
# Read fire data ----------------------
mtbs <- read_rds('../data/processed/mtbs.rds') %>%
  mutate(train = FIRE_YEAR < 2010)

if (!file.exists('max-plot.png')) {
  max_plot <- mtbs %>%
    group_by(FIRE_YEAR) %>%
    summarize(max_size = max(R_ACRES)) %>%
    ggplot(aes(FIRE_YEAR, max_size)) + 
    geom_point(data = mtbs, aes(y = R_ACRES), alpha = .01) +
    geom_point() + 
    scale_y_log10() + 
    xlab('Year') + 
    ylab('Fire size (acres)')
  ggsave('max-plot.png', plot = max_plot, width = 1.618 * h, height = h, dpi = 400)
}

```

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{max-plot}

\end{center}

# Two ideas

\huge{1. Changing size distributions}

# 

```{r}
len <- 100
density_df <- tibble(x = seq(-4, 5, length.out = len)) %>%
  mutate(old_den = dnorm(x), 
         new_den = dnorm(x, mean = 1), 
         exp_x = exp(x))

p <- density_df %>%
  ggplot(aes(x, old_den)) + 
  geom_line(alpha = .4) + 
  theme(axis.text = element_blank(),
        panel.grid = element_blank()) +
  xlab('Event size') + 
  ylab('Probability') + 
  annotate(x = -2, y = .3, label = '1984', alpha = .4, geom = 'text', 
           size = 6)
p
```

# 

```{r}
p + 
  geom_line(aes(y = new_den)) + 
  annotate(x = 3, y = .3, label = '2015', geom = 'text', 
           size = 6)
```


# Two ideas

1. Changing size distributions

\huge{2. More ignitions $\rightarrow$ more extremes}


#

```{r}
sampling_plot <- function(n = 5, seed) {
  set.seed(seed)
  samples <- tibble(x = rnorm(n))
  p <- density_df %>%
    ggplot(aes(x, old_den)) + 
    geom_line(alpha = .4) + 
    theme(axis.text = element_blank(),
          panel.grid = element_blank()) +
    xlab('Event size') + 
    ylab('Probability') +
    geom_vline(xintercept = max(samples$x), color = 'red') + 
    geom_rug(data = samples, inherit.aes = FALSE, aes(x), sides = 'b') + 
    ggtitle(paste('n =', n))
  p
}

seed <- 2
sampling_plot(n = 5, seed)
```

# 

```{r}
sampling_plot(n = 10, seed)
```

#

```{r}
sampling_plot(n = 50, seed)
```

#

```{r}
sampling_plot(n = 100, seed)
```


#

```{r}
sampling_plot(n = 1000, seed)
```

# 

```{r}
e_nb <- 2
s_nb <- 2

num_dist <- tibble(x = 0:30) %>%
  mutate(pr_y = dnbinom(x, mu = e_nb, size = s_nb))
p2 <- num_dist %>%
    ggplot(aes(x, pr_y)) + 
    geom_point() + 
    geom_linerange(aes(ymin = 0, ymax = pr_y)) +
    theme(axis.text = element_blank(),
          panel.grid = element_blank()) +
    xlab('Number of events') + 
    ylab('')

p2 + 
  ylab('Probability')
```

# 

```{r}
p1 <- density_df %>%
    ggplot(aes(x, old_den)) + 
    geom_line() + 
    theme(axis.text = element_blank(),
          panel.grid = element_blank()) +
    xlab('Event size') + 
    ylab('')


arrow_df <- data.frame(x = c(0),
                       xend = c(1),
                       y = c(0), 
                       yend = c(0))

p3 <- ggplot(arrow_df, aes(x, y)) + 
  geom_segment(aes(xend = xend, yend = yend), 
               arrow = arrow()) + 
  theme_void()

n_count <- 100
count_vals <- rnbinom(n_count, mu = e_nb, size = s_nb)
p4 <- expand.grid(n = count_vals,
            x = density_df$x) %>%
  as_tibble %>%
  mutate(id = rep(1:n_count, times = nrow(density_df)), 
         max_den = n * pnorm(x)^(n - 1) * dnorm(x)) %>%
  ggplot(aes(x, max_den, group = id)) + 
  geom_line(alpha = .1) + 
  xlab('Maximum size') + 
  ylab('') + 
  theme(axis.text = element_blank(),
      panel.grid = element_blank())

(p1 + p2 + plot_layout(ncol = 1)) | p3 | p4
```

# Our approach

### Model size and counts by:

- meteorological conditions
- housing density
- ecological context

**Can we predict extremes?**

# The data

1984-2009 (train), 2010-2015 (test)

```{r}
ecoregions <- read_rds('../data/processed/ecoregions.rds')

if (!file.exists('very-simple-ecoregions.shp')) {
  system("ogr2ogr -progress -simplify 200 very-simple-ecoregions.shp ../data/raw/us_eco_l3/us_eco_l3.shp")
}

simple_ecoregions <- read_sf('very-simple-ecoregions.shp')

if (!file.exists('fire-map.png')) {
  fire_map <- simple_ecoregions %>%
    ggplot() +
    geom_sf(size = .5, fill = 'white') + 
    geom_sf(data = mtbs, size = .1,
            inherit.aes = FALSE, color = 'red', alpha = .2) +
    theme_minimal() + 
    theme(panel.grid.major = element_line(colour = "lightgrey"), 
          legend.position = 'none', 
          axis.text = element_blank()) + 
    ggtitle('')
  ggsave('fire-map.png', plot = fire_map, width = 1.618 * h, height = h, dpi = 400)
}
```

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{fire-map}

\end{center}

# Bayesian spatiotemporal models

- Spatial splines
- Time-evolving Markov random fields


# Nonlinear effects

```{r, message = FALSE}

m <- 5
pal <- 2
axis_alpha <- .2

# Simulating function values ----------------------------------------------
n <- 100
x <- seq(0, 1, length.out = n)
unweighted_basis <- splines::bs(x, df = 5, intercept = TRUE) %>%
  as.data.frame() %>%
  as_tibble %>%
  mutate(x = x) %>%
  gather(key = 'dim', 'basis_value', -x) %>%
  mutate(dim = as.integer(dim), 
         weight = 1,
         type = 'Unweighted')

generate_f <- function(weights) {
  stopifnot(length(weights) == m)
  unweighted_basis %>%
  mutate(type = 'Weighted', 
         weight = weights[dim], 
         weighted_value = weight * basis_value) %>%
  group_by(x) %>%
  summarize(f_x = sum(weighted_value))
}

set.seed(123456)
er1_w <- rnorm(m)
er2_w <- er1_w + rnorm(m, mean = 0, sd = .2)
er3_w <- rnorm(m)

f1 <- generate_f(er1_w) %>%
  mutate(NA_L3NAME = "Central Rockies")
f1 %>%
  ggplot(aes(x, f_x, color = NA_L3NAME)) + 
  geom_line() + 
  xlab('Explanatory variable') + 
  ylab('Effect') + 
  theme(axis.text = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = 'none') + 
  scale_color_gdocs()
```

# Nonlinear effects

```{r, message = FALSE}
f2 <- generate_f(er2_w) %>%
  mutate(NA_L3NAME = 'Southern Rockies')

f1 %>%
  full_join(f2) %>%
  ggplot(aes(x, f_x, color = NA_L3NAME)) + 
  geom_line() + 
  xlab('Explanatory variable') + 
  ylab('Effect') + 
  theme(axis.text = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = 'none') + 
  scale_color_gdocs()
```

# Nonlinear effects

```{r, message = FALSE}
f3 <- generate_f(er3_w) %>%
  mutate(NA_L3NAME = 'Z region')

f1 %>%
  full_join(f2) %>%
  full_join(f3) %>%
  ggplot(aes(x, f_x, color = NA_L3NAME)) + 
  geom_line() + 
  xlab('Explanatory variable') + 
  ylab('Effect') + 
  theme(axis.text = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = 'none') + 
  scale_color_gdocs()
```

# 

\Huge{Model comparisons}

# Burn area model comparisons

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/ppc-density-funs}

\end{center}


# Count model comparisons

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/ppc-counts}

\end{center}


#

\Huge{What we found: fire size}

# Dry conditions $\rightarrow$ larger fires

```{r}
plot_data <- read_rds('../data/processed/plot_data.rds')
burn_area_partials <- read_rds("../data/processed/burn_area_partials.rds")
st_covs <- read_rds('../data/processed/st_covs.rds')
min_size <- 1e3

humidity_area_plot <- burn_area_partials %>%
  ggplot(aes(rmin, med, group = NA_L3NAME)) + 
  geom_line(alpha = .1) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  xlab('Humidity') + 
  ylab('Effect on fire size')
humidity_area_plot
```

92\% interval coverage

# Arizona/New Mexico Mountains

```{r, fig.width = 6, fig.height = 4}
plot_jpeg = function(url) {
  jpg <- RCurl::getURLContent(url) %>%
    jpeg::readJPEG(native = T)
  res <- dim(jpg)[2:1] # get the resolution, [x, y]
  plot(1, 1, xlim=c(1,res[1]), ylim=c(1,res[2]),
       asp = 1, type = 'n', xaxs = 'i', yaxs = 'i',
       xaxt = 'n', yaxt = 'n', xlab = '', ylab = '', bty = 'n')
  rasterImage(jpg,1,1,res[1],res[2])
}

plot_jpeg('https://upload.wikimedia.org/wikipedia/commons/thumb/f/f4/SUPERSTITIONS_AZ15.jpg/600px-SUPERSTITIONS_AZ15.jpg')
```

# 

```{r, fig.height = 3}
loc_ts <- read_rds('../data/processed/loc_ts.rds')

er_hum_ts <- function(ecoregion) {
  ts <- loc_ts %>%
    filter(NA_L3NAME == ecoregion) %>%
    mutate(train = ym < 2010)
  
  p1 <- ts %>%
    ggplot(aes(ym, rmin)) + 
    geom_line() + 
    xlab('') + 
    ylab('Humidity') + 
    ylim(5, 60)
  
  p2 <- ts %>%
    ggplot(aes(ym, exp(median) + min_size, fill = train)) + 
    geom_ribbon(aes(ymin = exp(lo) + min_size, ymax = exp(hi) + min_size), 
                alpha = .4) +
    geom_line() + 
    xlab('') + 
    ylab('E(acres)') + 
    theme(legend.position = 'none')
  
  p1 / p2  
}

er_hum_ts("Arizona/New Mexico Mountains")
```

# Atlantic Coastal Pine Barrens

```{r, fig.width = 6, fig.height = 4}
plot_jpeg("https://upload.wikimedia.org/wikipedia/commons/thumb/6/6d/Mullica_River_5.jpg/640px-Mullica_River_5.jpg")
```

# 

```{r, fig.height = 3}
er_hum_ts('Atlantic Coastal Pine Barrens')
```

#

\Huge{What we found: fire counts}

# Dry conditions $\rightarrow$ more fires

```{r}
humidity_count_plot <- plot_data %>%
  filter(covariate == 'rmin') %>%
  ggplot(aes(covariate_value, med, group = NA_L3NAME)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  geom_line(alpha = .1) + 
  xlab('Humidity') + 
  ylab('Effect on number of fires')
humidity_count_plot
```

99\% interval coverage

# Temperature effects are context dependent

#

```{r}
temp_count_plot <- plot_data %>%
  filter(covariate == 'tmmx') %>%
  ggplot(aes(covariate_value, med, group = NA_L3NAME)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  geom_line(alpha = .1) + 
  xlab('Temperature (C)') + 
  ylab('Effect on number of fires')
temp_count_plot
```

# Great Plains

```{r}
plot_data %>%
  filter(covariate == 'tmmx') %>%
  mutate(highlight = NA_L1NAME == 'GREAT PLAINS') %>%
  ggplot(aes(covariate_value, med, group = NA_L3NAME, 
             color = highlight)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  geom_line() + 
  xlab('Temperature (C)') + 
  ylab('Effect on number of fires') + 
  theme(legend.position = 'none') +
  scale_color_manual(values = c(scales::alpha(1, .1), 
                                'dodgerblue'))
```

# Ozark/Ouachita-Appalachian Forests

```{r}
plot_data %>%
  filter(covariate == 'tmmx') %>%
  mutate(highlight = NA_L2NAME == 'OZARK/OUACHITA-APPALACHIAN FORESTS') %>%
  ggplot(aes(covariate_value, med, group = NA_L3NAME, 
             color = highlight)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  geom_line() + 
  xlab('Temperature (C)') + 
  ylab('Effect on number of fires') + 
  theme(legend.position = 'none') +
  scale_color_manual(values = c(scales::alpha(1, .1), 
                                'dodgerblue'))
```

# Northwestern forested mountains

```{r}
plot_data %>%
  filter(covariate == 'tmmx') %>%
  mutate(highlight = NA_L1NAME == 'NORTHWESTERN FORESTED MOUNTAINS') %>%
  ggplot(aes(covariate_value, med, group = NA_L3NAME, 
             color = highlight)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  geom_line() + 
  xlab('Temperature (C)') + 
  ylab('Effect on number of fires') + 
  theme(legend.position = 'none') +
  scale_color_manual(values = c(scales::alpha(1, .1), 
                                'dodgerblue'))
```


# Putting it all together

```{r}
(p1 + p2 + plot_layout(ncol = 1)) | p3 | p4
```

# 

```{r, fig.width = 4, fig.height = 3, warning=FALSE}
interval_df <- read_csv('../data/processed/mev_intervals.csv')
to_plot <- c("Arizona/New Mexico Mountains", 
             "Atlantic Coastal Pine Barrens")
interval_df %>%
  filter(NA_L3NAME %in% to_plot) %>%
  group_by(NA_L3NAME) %>%
  mutate(total_n = sum(!is.na(empirical_max))) %>%
  ggplot(aes(x = ym, group = NA_L3NAME)) +
  geom_ribbon(aes(ymin = m_qlo, ymax = m_qhi),
              color = NA,
              fill = 'firebrick', alpha = .7) +
  geom_ribbon(aes(ymin = m_qvlo, ymax = m_qvhi),
              color = NA,
              fill = 'firebrick', alpha = .3) +
  scale_y_log10() +
  theme_minimal() +
  facet_wrap(~ fct_reorder(l3_er, rmin),
             labeller = labeller(.rows = label_wrap_gen(23)), 
             ncol = 1) +
  geom_point(aes(y = empirical_max), size = .5) +
  xlab('') +
  ylab('Maximum fire size (acres)') +
  theme(panel.grid.minor = element_blank(), 
        axis.text.x = element_text(angle = 90))
```

78\% interval coverage

# 

```{r, fig.width = 4.5}
case_study_plot <- read_rds('../data/processed/case-study-plot.rds')

date_df <- expand.grid(year = 2010:2015, 
                       month = 1:12) %>%
  as_tibble %>%
  mutate(ym = zoo::as.yearmon(paste(year, month, sep = '-')), 
         month_abb = month.abb[month])

case_study_plot$plot_data %>%
  filter(Response != "Zero-inflation component (number of fires)") %>%
  mutate(Response = ifelse(grepl("Lognormal", Response), "Fire size", "Fire count")) %>%
  left_join(date_df) %>%
  ggplot(aes(ym, y = median_eff, color = variable)) +
    geom_line() +
    facet_wrap(~Response, scales = 'free_y', ncol = 1) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank()) +
    xlab('') +
    ylab('Contribution') +
    scale_color_manual('', values = c('dodgerblue',
                                              'red',
                                              scales::alpha(
                                                c('green4',
                                                  'pink',
                                                  'orange',
                                                  'lightblue'), .7)))
```

# 

```{r, fig.width = 6, fig.height = 5}
cake_url <- "https://upload.wikimedia.org/wikipedia/commons/thumb/8/84/Chocolate_Cake_Flourless_%281%29.jpg/1024px-Chocolate_Cake_Flourless_%281%29.jpg"
plot_jpeg(cake_url)
```

# Million acre event probabilities

```{r, fig.width = 4.5, fig.height = 4, warning=FALSE}
mill_summ <- read_rds('../data/processed/mill_summ.rds')
map_plot <- simple_ecoregions %>%
  left_join(mill_summ) %>%
  ggplot(aes(fill = pr_million_acre_event)) +
  geom_sf(size = .1, color = 'white') + 
  theme_minimal() + 
  theme(panel.grid.major = element_line(colour = "lightgrey"), 
        axis.text = element_blank()) + 
  ggtitle('') + 
  scale_fill_viridis_c(trans = 'log', 
                       breaks = c(.001, .01, 
                                  .04, .07), 
                       "Probability", 
                       option = 'C')
point_plot <- mill_summ %>%
  arrange(-pr_million_acre_event) %>%
  top_n(14) %>%
  mutate(NA_L3NAME = gsub(" and ", " & ", NA_L3NAME)) %>%
  ggplot(aes(x = pr_million_acre_event, y = reorder(NA_L3NAME, pr_million_acre_event))) + 
  geom_point() + 
  geom_segment(aes(x = 0, xend = pr_million_acre_event, yend = reorder(NA_L3NAME, pr_million_acre_event))) + 
  ylab("") + 
  xlab("Probability")

map_plot
```

Overall: $\approx$ 35\% chance

# Million acre event probabilities

```{r, fig.width = 4.5}
point_plot
```

# 

# 

\Huge{Extreme fire risk is doubly nonlinear}

# 

```{r, fig.height = 3.6}
humidity_area_plot / humidity_count_plot
```

#

\Huge{Temperature effects vary}

# 

```{r}
plot_data %>%
  filter(covariate == 'tmmx') %>%
  mutate(highlight = NA_L1NAME == 'NORTHWESTERN FORESTED MOUNTAINS') %>%
  ggplot(aes(covariate_value, med, group = NA_L3NAME, 
             color = highlight)) + 
  geom_hline(yintercept = 0, linetype = 'dashed') + 
  geom_line() + 
  xlab('Temperature (C)') + 
  ylab('Effect on number of fires') + 
  theme(legend.position = 'none') +
  scale_color_manual(values = c(scales::alpha(1, .1), 
                                'dodgerblue'))
```

# 

\Huge{Extremes may be predictable...}

\Huge{\textit{and we can do better!}}

# 

```{r, warning=FALSE}

interval_df %>%
  filter(NA_L3NAME == "Arizona/New Mexico Mountains") %>%
  group_by(NA_L3NAME) %>%
  mutate(total_n = sum(!is.na(empirical_max))) %>%
  ggplot(aes(x = ym, group = NA_L3NAME)) +
  geom_ribbon(aes(ymin = m_qlo, ymax = m_qhi),
              color = NA,
              fill = 'firebrick', alpha = .7) +
  geom_ribbon(aes(ymin = m_qvlo, ymax = m_qvhi),
              color = NA,
              fill = 'firebrick', alpha = .3) +
  scale_y_log10() +
  theme_minimal() +
  facet_wrap(~ fct_reorder(l3_er, rmin),
             labeller = labeller(.rows = label_wrap_gen(23)), 
             ncol = 1) +
  geom_point(aes(y = empirical_max), size = .5) +
  xlab('') +
  ylab('Maximum fire size (acres)') +
  theme(panel.grid.minor = element_blank(), 
        axis.text.x = element_text(angle = 90))
```

# 

**Check out the code:** github.com/mbjoseph/wildfire-extremes

**Get in touch:** maxwell.b.joseph@colorado.edu, @mxwlj


# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/spline-concept}

\end{center}


# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/maps}

\end{center}



# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/burn-area-effs}

\end{center}


# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/count-partial-effs}

\end{center}


# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/all-coefs}

\end{center}

# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/max-preds-l3}

\end{center}

# 

\begin{center}

\includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{../fig/attribution-plot}

\end{center}