---
# PLEASE SEE THE README for in depth description github.com/brentthorne/posterdown
#
#---POSTER SIZE & DEFAULT FONT---#
poster_height: "38in" # height in inches of poster
poster_width: "45in" # width in inches of poster
font_family: "helvet" # choose from typical latex fonts (example: "palatino")
font_size: "30pt" #please see github.com/brentthorne/posterdown for compatible options.

#---TITLE BOX OPTIONS---#
#ESSENTIALS
title: 'Predicting wildfire extremes with Bayesian finite sample maxima'
author: "Max Joseph^1^ Matt Rossi^1^ Nate Mietkiewicz^1^ Adam Mahood^1^ Megan Cattau^1^ Lise St. Denis^1^ Chelsea Nagy^1^ Virginia Iglesias^1^ John Abatzoglou^2^ Jennifer Balch^1^"
affiliation: "^1^Earth Lab, University of Colorado Boulder; ^2^Department of Geography, University of Idaho"
#STYLE & FORMATTING
titlebox_bgcol: "ffffff"  #Colour of the Title Box background
titlebox_bordercol: "ffffff" #Colour of the title Box border.
titlebox_shape: "all"
titlebox_borderwidth: "1cm"
title_textcol: "000000" #colour of title text
author_textcol: "000000" # Colour of author text
affiliation_textcol: "000000" # Colour of affiliation text
title_textsize: "huge"         # Poster title fontsize
author_textsize: "large"       # Author list font size
affiliation_textsize: "large"  # Affiliation font size
#ADDING LOGOS
logoleft_name: 'Figures/earthlab'
logoleft_width: '3in'
logoleft_xshift: '1in'
logoleft_yshift: '1in'
logoright_name: 'Figures/cireslogo'
logoright_width: '3in'
logoright_xshift: '-1in'
logoright_yshift: '1in'

#---POSTER BODY OPTIONS---#
body_bgcol: "ffffff" #colour of the poster main background
body_textsize: "normalsize"    # Size of the main poster body text
body_textcol: "000000" # Colour of main text in the body of poster
column_numbers: 3 # Number of columns that the poster has
column_margins: "1in" # Margin spacing for columns
columnline_col: "008080" #colour 
columnline_width: "0pt" #width of line between each column
#SECTION TITLE STYLING
sectitle_textcol: "ffffff" # Colour of the poster section titles
sectitle_bgcol: "1e90ff" # Colour of the section title box
sectitle_bordercol: "1e90ff" # Colour of the border around the section title box.
sectitle_borderwidth: "2mm" # Thicknes of the section title box border
sectitle_boxshape: "uphill" # Changes the shape of the section title box.

#---BIBLIOGRAPHY OPTIONS---#
bibliography: MyLibrary # name of the .bib file used for referencing
bibliography_spacing: 0.8 # sets the multiplier for line spacing of bibliography spacing (between 0 and 1)
bibliography_textsize: "small"  # size of the bibliography text size (handy for one too many references!)

#---OTHER---#
cite_col: "CC0000" #colour of ciation elements
url_col: "008080" #colour of url links
link_col: "008080" #colour of other links within the poster
footnote_textcol: "ffffff" # Colour of footnote text if used
header-includes: \renewcommand{\familydefault}{\sfdefault}
output: posterdown::posterdown_latex
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'asis',
                      fig.height = 2, fig.width = 4.5, out.width = "100%", 
                      warning = FALSE, fig.align = 'center', message = FALSE, 
                      dev='cairo_pdf')

library(tidyverse)
library(knitr)
library(kableExtra)
library(assertthat)
library(patchwork)

theme_set(theme_minimal() + 
            theme(panel.grid.minor = element_blank()))

count_test_intervals <- read_csv('../../data/processed/count_test_intervals.csv')

area_coverage <- read_csv('../../data/processed/area_coverage.csv') %>%
  filter(model == '../../ba_lognormal_fit.rds') %>%
  arrange(idx)

holdout_burns <- read_rds('../../data/processed/holdout_burns.rds') %>%
  as_tibble()

assert_that(all(area_coverage$NA_L3NAME == holdout_burns$NA_L3NAME))

burn_area_coverage <- holdout_burns %>%
  mutate(idx = 1:n()) %>%
  full_join(area_coverage)

hectares_per_acre <- 0.404686
```

<!--
Bellow you will find typical RMarkdown syntax which should fill in the columns for the poster from top to bottom left to right. The more information you have the more will be filled. Please let me know if you have feedback at brentthorne18@gmail.com or at github.com/brentthorne/posterdown.
ENJOY! XD
-->

# The largest wildfires are getting larger

```{r plot_maxima, message=FALSE, fig.height=2}
mtbs <- read_rds('../../data/processed/mtbs.rds') %>%
  mutate(train = FIRE_YEAR < 2010)

annual_maxima <- mtbs %>%
  group_by(FIRE_YEAR) %>%
  mutate(max_size = max(Acres), 
         max_hectares = max_size * hectares_per_acre) %>%
  filter(Acres == max_size)

max_plot <- annual_maxima %>%
  ggplot(aes(Ig_Date, max_hectares)) + 
  geom_point(data = mtbs, aes(y = Acres * hectares_per_acre), alpha = .01) +
  geom_point() + 
  scale_y_log10() + 
  xlab('Ignition date') + 
  ylab('Fire size (hectares)') + 
  scale_x_date(date_breaks = '5 years', date_labels = "%Y")
max_plot
```

# This might happen for two reasons

\large{\textbf{1. Changing distributions}}

Fire size distributions may be changing over time. 

```{r size_dist, fig.height = 2}
len <- 100
density_df <- tibble(x = seq(-4, 5, length.out = len)) %>%
  mutate(old_den = dnorm(x), 
         new_den = dnorm(x, mean = 1), 
         exp_x = exp(x))

p <- density_df %>%
  ggplot(aes(x, old_den)) + 
  geom_line(alpha = .4) + 
  theme(axis.text = element_blank(),
        panel.grid = element_blank()) +
  xlab('Event size') + 
  ylab('Probability') + 
  annotate(x = -2, y = .3, label = '1984', alpha = .4, geom = 'text', 
           size = 6)
p + 
  geom_line(aes(y = new_den)) + 
  annotate(x = 3, y = .3, label = '2016', geom = 'text', 
           size = 6)
```

\large{\textbf{2. Tail sampling}}

More ignitions make rare events more likely.

```{r sampling_dist}
sampling_plot <- function(n = 5, seed) {
  set.seed(seed)
  samples <- tibble(x = rnorm(n))
  p <- density_df %>%
    ggplot(aes(x, old_den)) + 
    geom_line(alpha = .4) + 
    theme(axis.text = element_blank(),
          panel.grid = element_blank()) +
    xlab('Event size') + 
    ylab('Probability') +
    geom_vline(xintercept = max(samples$x), color = 'red') + 
    geom_rug(data = samples, inherit.aes = FALSE, aes(x), sides = 'b') + 
    ggtitle(paste('n =', n))
  p
}

seed <- 2

p1 <- sampling_plot(n = 10, seed)
p2 <- sampling_plot(n = 100, seed)
p1 + p2
```

# Theoretical background

We extended the metastatistical extreme values approach [@marani2015metastatistical] to the spatiotemporal domain, and linked it to the Bayesian **posterior predictive distribution** of finite sample maxima. 
In this framework, all unknowns (including future predictions) are treated as random variables. 


# The data

We used wildfire data from the Monitoring Trends in Burn Severity project, modeled at the Environmental Protection Agency level 3 ecoregion level for each month since 1984 [@eidenshink20071145801].

```{r plot_map, fig.width = 7, fig.height = 10}
library(sf)
library(tidyverse)
library(RColorBrewer)
library(patchwork)

ecoregions <- read_rds('../../data/processed/ecoregions.rds')

if (!file.exists('very-simple-ecoregions.shp')) {
  system("ogr2ogr -progress -simplify 200 very-simple-ecoregions.shp ../../data/raw/us_eco_l3/us_eco_l3.shp")
}

simple_ecoregions <- read_sf('very-simple-ecoregions.shp')

simple_ecoregions %>%
  ggplot() +
  geom_sf(size = .5, fill = 'white') + 
  geom_sf(data = mtbs, size = .1,
          inherit.aes = FALSE, color = 'red', alpha = .2) +
  theme_minimal() + 
  theme(panel.grid.major = element_line(colour = "lightgrey"), 
        legend.position = 'none', 
        axis.text = element_blank()) + 
  ggtitle('')
```

# The model

We developed spatiotemporal Bayesian count and size models with **spatial splines** and **time evolving Markov random fields**.
We used climate and housing covariates, evaluating predictive performance with a train/test split. 


```{r conceptual-figure, fig.height = 3}
e_nb <- 2
s_nb <- 2

num_dist <- tibble(x = 0:30) %>%
  mutate(pr_y = dnbinom(x, mu = e_nb, size = s_nb))


p1 <- density_df %>%
    ggplot(aes(x, old_den)) + 
    geom_line() + 
    theme(axis.text = element_blank(),
          panel.grid = element_blank()) +
    xlab('Event size') + 
    ylab('')

p2 <- num_dist %>%
    ggplot(aes(x, pr_y)) + 
    geom_point() + 
    geom_linerange(aes(ymin = 0, ymax = pr_y)) +
    theme(axis.text = element_blank(),
          panel.grid = element_blank()) +
    xlab('Number of events') + 
    ylab('')


arrow_df <- data.frame(x = c(0),
                       xend = c(0),
                       y = c(1), 
                       yend = c(0))

p3 <- ggplot(arrow_df, aes(x, y)) + 
  geom_segment(aes(xend = xend, yend = yend), 
               arrow = arrow()) + 
  theme_void()

n_count <- 100
count_vals <- rnbinom(n_count, mu = e_nb, size = s_nb)
p4 <- expand.grid(n = count_vals,
            x = density_df$x) %>%
  as_tibble %>%
  mutate(id = rep(1:n_count, times = nrow(density_df)), 
         max_den = n * pnorm(x)^(n - 1) * dnorm(x)) %>%
  ggplot(aes(x, max_den, group = id)) + 
  geom_line(alpha = .1) + 
  xlab('Maximum size') + 
  ylab('') + 
  theme(axis.text = element_blank(),
      panel.grid = element_blank())

(p1 / p2) + p3 + p4
```

# The code

The entire analysis and manuscript is reproducible and bundled up in a Docker container. 
To reproduce or extend this work, please fork us on GitHub: 

https://github.com/mbjoseph/wildfire-extremes

# Most extreme wildfires are predictable

Fire counts were predicted with 99\% coverage, and fire sizes were predicted with 93\% coverage. 
**Extreme events** were harder to predict, but the model achieved **75\% coverage**. 

```{r, extreme-ribbons, fig.height=5, fig.width = 5}
interval_df <- read_csv('../../data/processed/mev_intervals.csv')
interval_df %>%
  group_by(NA_L3NAME) %>%
  mutate(total_n = sum(!is.na(empirical_max))) %>%
  filter(total_n > 26) %>%
  ggplot(aes(x = ym, group = NA_L3NAME)) +
  geom_ribbon(aes(ymin = m_qlo * hectares_per_acre, ymax = m_qhi * hectares_per_acre),
              color = NA,
              fill = 'firebrick', alpha = .7) +
  geom_ribbon(aes(ymin = m_qvlo * hectares_per_acre, ymax = m_qvhi * hectares_per_acre),
              color = NA,
              fill = 'firebrick', alpha = .3) +
  scale_y_log10() +
  theme_minimal() +
  facet_wrap(~ fct_reorder(l3_er, rmin),
             labeller = labeller(.rows = label_wrap_gen(23)), 
             nrow = 4) +
  geom_point(aes(y = empirical_max * hectares_per_acre), size = .5) +
  xlab('') +
  ylab('Maximum fire size (hectares)') +
  theme(panel.grid.minor = element_blank(), 
        axis.text.x = element_text(angle = 90))
```


# Some extremes are hard to predict

The model failed to predict the largest event in the held-out test period: 2011's Wallow Fire because of the discretization of space and time. 
Treating space and time as **continuous** may improve performance in some cases. 

# Find out more

This paper is currently accepted in Ecological Applications, but you can read our preprint today on bioRxiv by scanning this QR code with your phone. 

\begin{figure}[H]
 \centering
 \includegraphics[width=6cm,height=6cm,keepaspectratio]{Figures/qr-code}
\end{figure}

<!--- Here you can set the size of the citation text as well as remove the "References" section if you choose not to have one for some reason :) -->
\normalsize\printbibliography
